Building socket-aware BPF programs

### 2018-10-27 paper/paper.tex
paper: Flesh out verifier, api definition sections

    To build such awareness into a BPF helper, the verifier needs the ability
    to track the safety of the calls, including appropriate reference counting
    upon the underlying socket. This paper describes extensions to the verifier
    to perform tracking of references in a BPF program. This allows BPF
    developers to extend the UAPI with functions that allocate and release
    resources within the execution lifetime of a BPF program, and the verifier
    will validate that the resources are released exactly once prior to program
    completion.

@@

\begin{itemize}
\item BPF started with sockets, but only provided access to packets
\item Previous helper extensions allowed e.g. socket operation handling
\item Arbitrary socket lookup from socket/packet hooks allows more powerful BPF
      programs
\item Use cases revolve around building logic in socket/packet layer hooks that
      change the behaviour based on the existence and attributes of a socket.
\end{itemize}

@@

The BPF verifier must ensure that all BPF programs that are loaded into the
kernel are safe to execute. For accessing socket pointers, this requires
validation that memory accesses are within the bounds of the memory allocated
for the structure, and will not make any modifications to the socket fields
that would be incompatible with the core socket handling logic. This is
provided through extension of existing functionality which performs bounds
verification and offset conversion for programs operating on CGroups hooks.
Additionally, the underlying memory must remain associated with the socket for
the duration of accesses from the BPF program, which may be guaranteed by
taking a reference on sockets for the duration of their use. The following
sections explain how this is implemented through the introduction of a new
pointer type in the verifier, and through the extension of the verifier to
recognise assembly instructions that represent reference acquisition and
release.

@@

Sockets in Linux are reference-counted to track usage and handle memory
release. In general to ensure the safety of socket for the duration of access
in BPF programs, this reference count needs to be incremented when sockets are
referenced from BPF programs. Two options were considered for how to guarantee
this: Implicit reference tracking and explicit reference tracking.

\subsubsection{Implicit reference tracking.}

This calls for the BPF infrastructure to handle reference tracking, and to hide
this detail from the BPF API. Whenever a socket lookup function is called, a
reference is taken and the socket is added to a reference list. At the end of
the BPF program execution, the core code could walk this reference list and
release each reference. However, collecting references and releasing them at
the end of the BPF program invocation has the unfortunate overhead that even
the execution of BPF programs that do not make use of the new socket lookup
helpers would need to pay the cost of checking the list of references to free.
When dealing with sufficiently demanding use cases such as those which use XDP,
even the few instructions required to implement such a check may have a
measurable impact on performance.

\subsubsection{Explicit reference tracking.}

This calls for reference acquisition and release semantics to be built into the
API. If references must be taken, then requiring BPF program writers to
explicitly handle these references ensures that the program writer must
understand the potential impact this may have on the operation of the program
(including atomic operation cost), and it also ensures that the cost is only
borne by programs that make use of this feature.

\subsubsection{Implementation.}

Based upon the tradeoffs described above, explicit reference tracking was
chosen. From an implementation perspective, this requires tracking pointer
values through each conditional path in the program, and rejecting the load of
programs that fail to "balance" resource acquisition (lookup) with release.
This implementation works as follows.

Calls to helper functions which acquire references to resources are annotated
in the verifier to associate the acquisition and release functions. When the
instruction that acquires a resource is processed, a resource identifier is
allocated for this resource. This identifier is kept in a list in the verifier
state, and it is also associated with the register which receives the resulting
resource pointer. When processing a corresponding release helper call, if a
parameter to the function contains a register that is associated with a
resource identifier, then the identifier is removed from the verifier state. If
any paths in the program reach the final (\verb+BPF_EXIT+) instruction and the
verifier state contains any resource identifier association, then the program
is considered unsafe as it has leaked the resource.

Some instruction calls are restricted while holding a resource reference to
avoid leaking references, for example the \verb+bpf_tail_call+ helper function.

@@

The following considerations were made for implementing the lookup helpers. The
consideration of each of these items reflects the definition of the lookup
helpers in Listing~\ref{lst:helper-api}.

@@

@@

Multiple optimizations were proposed during development of this feature to
reduce the runtime overhead of using the socket lookup helpers. Two such
optimizations are described below: Skipping reference counting when unnecessary
to save on atomic instructions, and allowing the use of direct packet pointers
for the tuple.

@@

The implementation introduced in the 4.20 release cycle includes support for
looking up TCP and UDP sockets for IPv4 and IPv6 traffic, using the
\verb+SCHED_CLS+, \verb+SCHED_ACT+ and \verb+SK_SKB+ hook points. A proposed
patchset extends this to allow the same helpers to be used from the \verb+XDP+
hook points~\cite{sk-lookup-xdp}.

\todo{Associate custom metadata with socket?}

@@

Stateful firewalling is distinguished from stateless firewalling in that it
provides the ability to associate two directions of a connection together and
apply a filtering policy based upon the direction of the connection. This
commonly involves using a connection tracker to reconstruct the connection
state of each endpoint of the connection without specific knowledge from those
endpoints~\cite{ovs-ct,conntrack}. In cases where the Linux stack acts as a
firewall on a node in the network that is separate from both endpoints, In many
cases this model is used when filtering traffic where one endpoint is colocated
with the Linux stack that performs the firewalling, which already contains
information about the local end of the connection in a slightly.

>>>>

    This paper presents the changes necessary to extend socket introspection
    capabilities of the Linux kernel's BPF packet hooks. To provide safety
    guarantees around the access of the socket, the verifier was extended to
    track allocation and release of kernel resources. Using this mechanism, we
    introduce socket lookup and release helpers into the BPF API.

    %To provide access to socket state from a BPF helper call, the verifier
    %needs the ability to track the safety of the calls, including appropriate
    %reference counting upon the underlying socket for the duration of the
    %access.

    %With such an ability ability in the verifier, we add socket lookup and
    %release function calls to the BPF API, allowing BPF programs to safely find
    %a socket and build logic upon the presence or attributes of a socket.
    %Applications for this functionality range from load-balancing traffic based
    %on the presence of a listening application to building stateful firewalling
    %primitives to understand whether traffic for this connection has been seen
    %before. With this new functionality, BPF programs can integrate more
    %closely with the networking stack's understanding of the traffic transiting
    %the kernel.

    %provide safe access to sockets from BPF packet handling hooks. The BPF verifier
    %is extended to gain understanding of reference-counted pointer types, which
    %allows BPF code paths to

@@

% Over the past several years, BPF has steadily become more powerful in
% multiple ways: Through building more intelligence into the verifier which
% allows more complex programs to be loaded, and through extension of the API
% such as by adding new map types and new native BPF function calls. While
% BPF has its roots in applying filters at the socket layer, the ability to
% introspect the sockets relating to traffic being filtered has been limited.

To build such awareness into a BPF helper, the verifier needs the ability to
track the safety of the calls, including appropriate reference counting upon
the underlying socket. This talk walks through extensions to the verifier to
perform tracking of references in a BPF program. This allows BPF developers to
extend the UAPI with functions that allocate and release resources within the
execution lifetime of a BPF program, and the verifier will validate that the
resources are released exactly once prior to program completion.

\todo{Desire to target logic based upon socket existence/attributes at packet
layers}

@@

The BPF verifier must ensure that all BPF programs that are loaded into the
kernel are safe to execute. For accessing socket pointers, this requires
validation that memory accesses are within the bounds of the memory allocated
for the structure, and will not make any modifications to the socket fields
that would be incompatible with the core socket handling logic. This is
provided through extension of existing functionality which provides bounds
verification and offset conversion for programs operating on CGroups hooks.
Additionally, the underlying socket memory must remain associated with the
socket for the duration of accesses from the BPF program, which may be
guaranteed by taking a reference on sockets for the duration of their use.

@@

As with many other structure types, Linux sockets are reference-counted to
track usage and handle memory release. In general, to ensure the safety of
socket for the duration of access in BPF programs, when sockets are referenced
from BPF programs the reference count for the socket needs to be incremented.
Two options were considered for how to guarantee this: Implicit reference
tracking and explicit reference tracking.

Implicit reference tracking calls for the BPF infrastructure to handle
reference tracking, and to hide this detail from the BPF API. Whenever a socket
lookup function is called, a reference is taken and the socket is added to a
reference list. At the end of the BPF program execution, the core code could
walk this reference list and release each reference. However, collecting
references and releasing them at the end of the BPF program invocation has the
unfortunate overhead that even the execution of BPF programs that do not make
use of the new socket lookup helpers would need to pay the cost of checking the
list of references to free. When dealing with sufficiently demanding use cases
such as those which use XDP, even the few instructions required to implement
such a check may have a measurable impact on performance.

Explicit reference tracking calls for reference acquisition and release
semantics to be built into the API. If references must be taken, then requiring
BPF program writers to explicitly handle these references ensures that the
program writer must understand the potential impact this may have on the
operation of the program (including atomic operation cost), and it also ensures
that the cost is only borne by programs that make use of this feature.

\todo{Lead into the implementation} This safety is provided by tracking pointer
values through each conditional path in the program, and rejecting the load of
programs that fail to "balance" resource acquisition (lookup) with release.

Calls to helper functions which acquire references to resources are annotated
in the verifier to associate the acquisition and release functions. When the
instruction that acquires a resource is processed, the verifier state is
updated to include a resource identifier for the resulting value, and the
register's value is associated with this identifier. When processing a
corresponding release helper call, if a parameter to the function contains a
resource identifier, then the corresponding identifier is removed from the
verifier state. If any paths in the program reach the final (\verb+BPF_EXIT+)
instruction and the verifier state still contains resource identifiers, then
the program is considered unsafe as it has leaked the resource.

@@

The following considerations were made for implementing the lookup helpers:

@@

The consideration of each of these items reflects the definition of the lookup
helpers (Listing~\ref{lst:helper-api}):

@@

@@

The implementation intruduced in the 4.20 release cycle includes support for
looking up TCP and UDP sockets for IPv4 and IPv6 traffic, using the \verb+tc+
and \verb+sk_skb+ hook points. A proposed patchset extends this to allow the
same helpers to be used from the \verb+xdp+ hook points.

@@
